Bayesian Sparsification of $\cplx$-valued networks

### 2020-01-17 paper/paper.tex
cplx-network review, fixed an eqref

Signal processing, specifically, audio or radio frequency signal analysis is inherently
complex-valued, yet $\cplx$-valued models and neural networks, and, in particular, compression
methods tailored to $\cplx$-domain remain a niche field of research.

%%% >>>> (cplx)
% complex nets and related stuff
% cplx domain models
% (cplx-transformer) yang-ma-li.tex
% (cplx capacity) monning-manandhar.tex
\citet{monning_evaluation_2018} state,
that $\cplx$-networks have less intuitive architectures, inherently higher arithmetic
complexity due to multiplication in $\cplx$, that requires 3-4 floating point multiplications,
dependening on the implementation.
% due to them not being true cplx and just paired real computational graphs

\citep{yang_complex_2019,monning_evaluation_2018}

data that is naturally complex-valued
studies have shown a potentially richer representation of complex nets
\citep{hirose2003,fijulamin2009}
% [4] Md Faijul Amin and Kazuyuki Murase, “Single-layered complex-valued neural network for real-valued classification problems,” Neurocomputing, vol. 72, no. 4-6, pp. 945–955, 2009.

the complex version of backpropagation algorithm  was proposed because of
its richer representational capacity.
% Nevio Benvenuto and Francesco Piazza, “On the complex backpropagation algorithm,” IEEE Transactions on Signal Processing, vol. 40, no. 4, pp. 967–969, 1992.
that operates on the Fourier transform input speech, acoustic, of radio signals.

Application of complex-valued networks is not new, \citep{citation_needed}. More recently
they were applied to music transcription and speech recognition \citep{trabelsi_deep_2017},
and \citep{yang_complex_2019}. The key motivation behind adoption of $\cplx$-networks is
pervasiveness of $\cplx$-valued data in digital signal processing \citep{citation_needed},
acoustic and radio signal analysis \citep{citation_needed}, and quantum systems modelling
\citep{citation_needed}, etc. All these applications require compliance with algebraic
operations in the field of complex numbers.

\citep{yang_complex_2019} develop complex transformer, with $\cplx$-attention and complex
encoder-decoder, and apply it to music transcription and wireless signal classification.

papers related to complex-valued distributions \citep{pav_moments_2015,taubock_complex-valued_2012},
and \citep{karseras_caution:_2014}

%%% <<<< (cplx)

% motivation and other methods
Computational efficiency improvement and model compression methods such as quantization
\citep{uhlich_differentiable_2019}, integer-based arithmetic \citep{lin_fixed_2016,chen_fxpnet_2017},
and knowledge distillation \citep{hinton_distilling_2015} appear to be directly applicable
to $\cplx$-valued networks.
%
Certain model sparsification methods, e.g. sparsity inducing regularizers, and magnitude
pruning \citep{zuo_compression_2019}, which reactivates zeroed parameters based on gradients,
can be adapted directly to $\cplx$-valued models. Other methods require additional considerations.
% the more explicitly zero parameters there is, the less computations is needed.

@@

% our contribution
Inspired by variational sparsification and stochastic sparsification techniques we
develop bayesian dropout for $\cplx$-valued networks.
%
The paper is structured as follows. In sec.~\ref{sec:bayesian_dropout} we review Bayesian
dropout techniques, and in sec.~\ref{sec:c_valued_networks} we provide a brief summary of
the inner working of complex-valued networks as functions of complex argument. The main
contribution of this study is presented in sec.~\ref{sec:dropout_for_c_valued_layers},
where we consider different variational approximations and priors, outline the tricks
and derive expressions for the divergence term. In sec.~\ref{sec:experiments} we evaluate
the sparsification rate and compare the resulting performance of various $\cplx$-networks
proposed in prior work, and discuss the outcomes in sec.~\ref{sec:discussion_and_related_research}.

@@

Being the ratio of the squared mean to the effective variance, $\alpha_j$ regulates the
relevance of the parameter $\omega_j$ it is associated to. Thus \citet{molchanov_variational_2017}
focus on the model sparsification capabilities of Variational Dropout, optimizing $\alpha$
for each individual parameter. Flexibility in specification of $q_\theta$ enables structured
dropout with shared $\alpha$ across groups of parameters $\omega$. Another variant of Bayesian
dropout, called \textit{Automatic Relevance Determination}, is developed in \citep{kharitonov_variational_2018}.
The key difference is that the log-uniform factor $\pi(\omega_{ij})$ is replaced by a proper
Gaussian prior $
  \mathcal{N}(0, \tau^{-1}_{ij})
$ with learnable precision $\tau_{ij} > 0$. This method, known as \textit{Empirical Bayes}
\citep{citation_needed}, fits a prior distribution while performing variational inference.
Maximizing \eqref{eq:elbo} over $\lambda$, holding other parameters fixed, yields $
  \tau^* = {(\mu^2 + \sigma^2)}^{-1}
$, whence
\begin{equation}  \label{eq:ard-kl-div-real}
  KL(q_\theta \| \pi)
    = \log{\bigl(1 + \tfrac1{\alpha} \bigr)}
    \,.
\end{equation}

@@

\citet{trabelsi_deep_2017} outline the building blocks for deep $\cplx$-valued networks
and describe suitable representation and operations including convolutional and dense layers,
$\cplx$-valued activations, complex batch-normalization and suitable weight initialization.
Their approach makes a complex-valued network into an intricately connected $\real$-valued
computational graph that respects $\cplx$-arithmetic.
%
In particular, natural identification of $\cplx$ and $\real^2$ implies that linear layers $
  L \colon \cplx^m \to \cplx^n
$ act upon their inputs as follows:
\begin{equation}  \label{eq:cplx-lin-op}
  {(\real^n)}^{2}
    \to {(\real^m)}^{2}
    \colon (u, v)
      \mapsto \Bigl(
        P u - Q v,
        P v + Q u
      \Bigr)
    \,,
\end{equation}
where $
  P, Q \colon \real^{n} \to \real^{m}
$ are unique operators such that $L = P + j Q$. A $\cplx$-convolutional layer with kernel
$W$ can be implemented as two $\real$-convolutions with kernels $\Re{W}$ and $\Im{W}$ using
\eqref{eq:cplx-lin-op}. Algebraic operations, trigonometric functions, are implemented
through operations in $\real^2$, \citep{hirose_complex-valued_2009}. Activations, considered
in \citep{trabelsi_deep_2017} include independent componentwise application of $\real$-valued
non-linearities, $
  z = u + j v
    \mapsto \sigma(u) + j \sigma(v)
$, $\cplx$-differentiable exponential and trigonometric activations, and \textit{modRelu}
activation, which applies soft-thresholding to the modulus and preserves the angle: $
  z = r e^{j \phi}
    \mapsto e^{j \phi} (r - b)_+
  % z \mapsto z \bigl(
  %     1 - \tfracb{\lvert z \rvert}
  %   \bigr)_+
$ with a learnable threshold parameter $b \in \real$.

@@

If we assume a mean field $\cplx$-Gaussian approximate posteriror for $
  W \in \cplx^{n\times m}
$, then
\begin{equation}  \label{eq:c-gauss-vi-general-vec}
  \vec{W}
    \sim \mathcal{N}^{\cplx}_{nm} \bigl(
      \vec{\mu}, \diag{\vec{\Sigma}}, \diag{\vec{C}}
    \bigr)
  \,,
\end{equation}
where mean, \textit{variance} and relation are given by matrices $
  \mu, \Sigma, C \in \cplx^{n\times m}
$ with $\Sigma_{ij} \geq 0$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$. For any $x \in \cplx^m$ and $b\in \cplx^n$ $
  y = W x + b
    = (I_n \otimes x^\top) \vec{W} + b
$, whence the covariance and relation matrices of $y$ are
\begin{align}
  (I \otimes x^\top) \diag{\vec{\Sigma}} (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (I \otimes x^\top) e_i \otimes e_j
    %   \Sigma_{ij} (e_i \otimes e_j)^\top (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (e_i \otimes x^\top e_j)
    %   \Sigma_{ij} (e_i \otimes x^\top e_j)^{\hop}
    % = \sum_{ij} (e_i \otimes x_j) \Sigma_{ij} (e_i^\top \otimes \conj{x}_j)
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m \Sigma_{ij} x_j \conj{x}_j
    \,,  \label{eq:cn-gauss-lrt-cov} \\
  (I \otimes x^\top) \diag{\vec{C}} (I \otimes x^\top)^\top
    % = \sum_{ij} (e_i \otimes x_j) C_{ij} (e_i \otimes x_j)^\top
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m C_{ij} x_j^2
    \,.  \label{eq:cn-gauss-lrt-rel}
\end{align}
Since \eqref{eq:cn-gauss-lrt-cov} and \eqref{eq:cn-gauss-lrt-rel} are diagonal, $(y_i)_{i=1}^n$
are independent whence \eqref{eq:cn-affine} implies
\begin{equation}  \label{eq:cplx-gauss-trick-appendix}
  y_i
    \sim \mathcal{N}^{\cplx}
      \bigl(
        e_i^\top \mu x + b_i,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \bigr)
    \,.
\end{equation}
Hence \eqref{eq:cplx-gauss-trick-appendix} folows.

>>>>

Inherently complex-valued nature of data in signal processing, specifically, acoustic or radio
frequency signal analysis, has been the key driver behind adoption of $\cplx$-networks.
%
\citet{hirose_complex-valued_2009} argues that the key merit of $\cplx$-valued neural
networks is simultaneous phase rotation and amplitude adjustment, intrinsic to multiplication
in $\cplx$ field. Despite inherently higher arithmetic complexity, which stems from up to four
floating point operations per single $\cplx$ multiplication, \citep{monning_evaluation_2018},
$\cplx$-arithmetic reduces ineffective degrees of freedom in comparison with an analogous
double-dimensional $\real$-valued network. \citet{hirose_complex-valued_2009} list numerous
applications related to acoustic and radio signal processing and demonstrate superiority of
$\cplx$-networks in the task of shallow landmine detection using ground penetrating radar
imaging. More recently they were applied to music transcription and speech recognition
\citep{trabelsi_deep_2017}. \citet{arjovsky_unitary_2016} propose $\cplx$-valued unitary-RNNs
to deal with exploding~/~vanishing gradients and apply the recurrent models to long-term
dependence prediction tasks. \citet{wisdom_full-capacity_2016} expand on their work by optimizing
recurrent weights on the $\cplx^{n \times n}$ Stiefel manifold. And \citet{wolter_complex_2018}
investigate different $\cplx$-valued gating mechanisms unitary-RNNs and conduct experiments
on human motion prediction and music transcription. \citet{yang_complex_2019} develop complex
transformer, with $\cplx$-attention and complex encoder-decoder, and apply it to music
transcription and wireless signal classification.

% complex-valued distributions \citep{pav_moments_2015,taubock_complex-valued_2012},
% and \citep{karseras_caution:_2014}

% motivation and other methods
However, in spite of relative success of $\cplx$-valued networks, compression methods tailored
to $\cplx$-domain remain a niche field of research.
%
Computational efficiency improvement and model compression methods such as quantization
\citep{uhlich_differentiable_2019}, integer-based arithmetic \citep{lin_fixed_2016,chen_fxpnet_2017},
and knowledge distillation \citep{hinton_distilling_2015} appear to be directly applicable
to $\cplx$-valued networks.
%
Certain model sparsification methods, e.g. sparsity inducing regularizers, and magnitude
pruning \citep{zuo_compression_2019}, which reactivates zeroed parameters based on gradients,
can be adapted directly to $\cplx$-valued models. Other methods require additional considerations.
% the more explicitly zero parameters there is, the less computations is needed.

@@

% our contribution
Inspired by variational sparsification and stochastic sparsification techniques we
develop bayesian dropout for $\cplx$-valued networks.
%
The paper is structured as follows. In sec.~\ref{sec:bayesian_dropout} we review Bayesian
dropout techniques, and in sec.~\ref{sec:c_valued_networks} we provide a brief summary of
the inner working of complex-valued networks as functions of complex argument. The main
contribution of this study is presented in sec.~\ref{sec:dropout_for_c_valued_layers},
where we consider different variational approximations and priors, outline the tricks
and derive expressions for the divergence term. In sec.~\ref{sec:experiments} we evaluate
the sparsification rate and compare the resulting performance of various $\cplx$-networks
proposed in prior work, and discuss the outcomes.

@@

Being the ratio of the squared mean to the effective variance, $\alpha_j$ regulates the
relevance of the parameter $\omega_j$ it is associated to. Thus \citet{molchanov_variational_2017}
focus on the model sparsification capabilities of Variational Dropout, optimizing $\alpha$
for each individual parameter. Flexibility in specification of $q_\theta$ enables structured
dropout with shared $\alpha$ across groups of parameters $\omega$. Another variant of Bayesian
dropout, called \textit{Automatic Relevance Determination}, is developed in \citep{kharitonov_variational_2018}.
The key difference is that the log-uniform factor $\pi(\omega_{ij})$ is replaced by a proper
Gaussian prior $
  \mathcal{N}(0, \tau^{-1}_{ij})
$ with learnable precision $\tau_{ij} > 0$. This method, known as \textit{Empirical Bayes}
\citep{citation_needed}, fits a prior distribution while performing variational inference.
Maximizing \eqref{eq:elbo} over $\tau$, holding other parameters fixed, yields $
  \tau^* = {(\mu^2 + \sigma^2)}^{-1}
$, whence
\begin{equation}  \label{eq:ard-kl-div-real}
  KL(q_\theta \| \pi)
    = \log{\bigl(1 + \tfrac1{\alpha} \bigr)}
    \,.
\end{equation}

@@

\citet{trabelsi_deep_2017} outline the building blocks for deep $\cplx$-valued networks
and describe suitable representation and operations including convolutional and dense layers,
$\cplx$-valued activations, complex batch-normalization and suitable weight initialization.
Their approach makes a complex-valued network into an intricately connected $\real$-valued
computational graph that respects $\cplx$-arithmetic.
%
In particular, natural identification of $\cplx$ and $\real^2$ implies that linear layers $
  L \colon \cplx^m \to \cplx^n
$ act upon their inputs as follows:
\begin{equation}  \label{eq:cplx-lin-op}
  {(\real^n)}^{2}
    \to {(\real^m)}^{2}
    \colon (u, v)
      \mapsto \Bigl(
        P u - Q v,
        P v + Q u
      \Bigr)
    \,,
\end{equation}
where $
  P, Q \colon \real^{n} \to \real^{m}
$ are unique operators such that $L = P + j Q$. A $\cplx$-convolutional layer with kernel
$W$ can be implemented as two $\real$-convolutions with kernels $\Re{W}$ and $\Im{W}$ using
\eqref{eq:cplx-lin-op}. Activations include trigonometric functions and representation-dependent
non-linearities, \citep{hirose_complex-valued_2009}: polar (phase-amplitude) $
  z \mapsto \sigma(\lvert z \rvert) e^{j \arg z}
$ or planar (real-imaginary) $
  z \mapsto \sigma(\Re z) + j \sigma(\Im z)
$ for some $\real$-valued non-linearity $\sigma$, possibly with learnable parameters,
\citep{trabelsi_deep_2017,wolter_complex_2018}.

@@

If we assume a mean field $\cplx$-Gaussian approximate posteriror for $
  W \in \cplx^{n\times m}
$, then
\begin{equation}  \label{eq:c-gauss-vi-general-vec}
  \vec{W}
    \sim \mathcal{N}^{\cplx}_{nm} \bigl(
      \vec{\mu}, \diag{\vec{\Sigma}}, \diag{\vec{C}}
    \bigr)
  \,,
\end{equation}
where mean, \textit{variance} and relation are given by matrices $
  \mu, \Sigma, C \in \cplx^{n\times m}
$ with $\Sigma_{ij} \geq 0$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$. For any $x \in \cplx^m$ and $b\in \cplx^n$ $
  y = W x + b
    = (I_n \otimes x^\top) \vec{W} + b
$, whence the covariance and relation matrices of $y$ are
\begin{align}
  (I \otimes x^\top) \diag{\vec{\Sigma}} (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (I \otimes x^\top) e_i \otimes e_j
    %   \Sigma_{ij} (e_i \otimes e_j)^\top (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (e_i \otimes x^\top e_j)
    %   \Sigma_{ij} (e_i \otimes x^\top e_j)^{\hop}
    % = \sum_{ij} (e_i \otimes x_j) \Sigma_{ij} (e_i^\top \otimes \conj{x}_j)
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m \Sigma_{ij} x_j \conj{x}_j
    \,,  \label{eq:cn-gauss-lrt-cov} \\
  (I \otimes x^\top) \diag{\vec{C}} (I \otimes x^\top)^\top
    % = \sum_{ij} (e_i \otimes x_j) C_{ij} (e_i \otimes x_j)^\top
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m C_{ij} x_j^2
    \,.  \label{eq:cn-gauss-lrt-rel}
\end{align}
Since \eqref{eq:cn-gauss-lrt-cov} and \eqref{eq:cn-gauss-lrt-rel} are diagonal, $(y_i)_{i=1}^n$
are independent whence \eqref{eq:cn-affine} implies
\begin{equation}  \label{eq:cplx-gauss-trick-appendix}
  y_i
    \sim \mathcal{N}^{\cplx}
      \bigl(
        e_i^\top \mu x + b_i,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \bigr)
    \,.
\end{equation}
Hence \eqref{eq:cplx-gauss-trick} folows.
